Classification Techniques for Conformance and Performance Checking in Process Analysis

### 2013-11-09 paper/Paper_aibp/camera-ready-paper2.tex
camera-ready copy added

\documentclass{llncs}
\usepackage[latin1,utf8x]{inputenc}
\usepackage{color}
\usepackage[pdftex]{graphicx}
\usepackage{amsmath}
\usepackage{verbatim}
\usepackage{marginnote} 
%\usepackage[outer=5cm, inner=2cm, heightrounded, marginparwidth=4cm, marginparsep=0.5cm]{geometry}

@@

\pagestyle{plain}

\title{Classification Techniques for Conformance and Performance Checking in Process Analysis}
\author{Hind Chfouka\inst{1}, Andrea Corradini\inst{1} and Roberto Guanciale\inst{2}}
\institute{Department of Computer Science, University of Pisa, Italy\and School of Computer Science and Communication,
KTH Royal Institute of Technology, Stockholm, Sweden}
\date{today}

@@

%% Classification techniques in the Process Mining field have already been explored~\cite{DBLP:conf/bpm/RozinatA06}\cite{DBLP:conf/sac/LeoniA13}. In~\cite{DBLP:conf/bpm/RozinatA06} the idea is exploring how data influences the case routing of process flow execution. In this contribution the goal is to strengthen the use of data analysis techniques for the benefit of process analysis. This is done by studying how data may influence, in general, the overall process behavior.

@@

\section{The case study: a sale business process}\label{example}
A business process can be seen as a collection of activities occurring within an organization that lead to a specific goal. There are several formalisms for representing business processes. In this paper we represent them using Petri nets~\cite{DBLP:conf/ifip/Petri62,DBLP:journals/jcsc/Aalst98} since this makes possible to exploit some existing process analysis techniques that are based on such nets. However, in a business management context, Petri nets can be not expressive enough and usually process models are presented as workflows, for example using a standard notation for the business process modeling: BPMN (Business Process Model and Notation)~\cite{BPMN}. Transforming a BPMN model into a Petri net is possible thanks to a mapping technique presented in~\cite{DBLP:conf/wsfm/BruniCFFGS11}.
\begin{figure}[h]
\includegraphics[width=360pt]
{./items/Sales_PN.pdf}
\caption{Petri net for a sale process}
\label{pnet}
\end{figure}
%
Figure~\ref{pnet} presents the Petri net model for a sale
process.\footnote{\small{
The Petri net of Figure~\ref{pnet} is obtained by
transforming a BPMN model using the algorithm of~\cite{DBLP:conf/wsfm/BruniCFFGS11}, and mapping
each activity to a pair of start/end transitions. The resulting net
has been simplified by removing the unnecessary \emph{invisible
  transitions} used to encode the \emph{Join} or \emph{Fork} gateway~\cite{DBLP:conf/wsfm/BruniCFFGS11}.
}} 
This business process abstractly represents a fragment of the
procedure followed in a commercial organization for managing orders
received from clients. In general this procedure includes several
activities and each of them involves a specific department of the
organization. We assume that the sale process begins with the
notification of an order from a client, so the first activity is
simply called \emph{Order}. After the order is received, the sale
process continues with some activities that can be done in parallel,
since they do not present any dependency. The \emph{FinancialCheck}
activity represents the financial analysis of the order  (i.e
verifying the financial situation of the client). Concurrently, the
\emph{WarehouseCheck} activity checks the availability of merchandise
requested by the order and, in case of a shortage, starts a supplying procedure. This is done through the activity \emph{ExternalWarehouse}. After the completion of the  parallel activities, a synchronization is needed before continuing the sale procedure with the final activity called \emph{Notify}, by which a result regarding the acceptance of the order is communicated to the client.

%Note that the Petri net of Figure~\ref{pnet} is obtained by transforming a BPMN model using the algorithm described in~\cite{DBLP:conf/wsfm/BruniCFFGS11}, and mapping each activity to a pair of start/end transitions. The resulting net has been simplified by removing the unnecessary \emph{invisible transitions} used to encode e.g. the \emph{Join} or \emph{Fork} gateway~\cite{DBLP:conf/wsfm/BruniCFFGS11}.

\section{Process Analysis}\label{Background}
%Process analysis is performed using the Petri net representing the process and an event log recording the related executions, also called \emph{process instances}. The basic building blocks of event logs are events. An event {\itshape e} can be seen as a pair {\itshape e = (a,t) } representing an action {\itshape a } recorded and the corresponding timestamp {\itshape t}. Action and timestamp are denoted respectively by $\alpha(e)$ and $\phi(e)$. Events that belong to the same process are grouped into {\itshape traces}. Formally, a trace $T$ is a finite sequence of events $T[1],..., T[n]$, such that $\phi(T[i]) \leq \phi(T[i+1])$ for all  $i \in [1,n)$. A log $L$ is a set of traces, recording the activities performed by a system during a finite number of process executions.
Process analysis is performed using the Petri net representing the process and an event log recording the related executions, also called \emph{process instances}. The basic building blocks of event logs are \emph{events}, that record the execution of the process actions. Therefore an event $e$ can be seen as a tuple including the action recorded, a corresponding timestamp, and possibly other attributes describing relevant information about the state of the process.  
%in gen pair $e = (a,t)$ representing an action $a$ recorded and the corresponding timestamp $t$.  
Events that belong to the same process instance are grouped in a {\itshape trace}, where events are ordered 
% A trace $T$ is a finite sequence of events ordered 
by timestamp. An \emph{event log} is a set of traces, recording the activities performed by a system during a finite number of process executions.  In this paper the following assumptions are made:
\begin{itemize}
\item All traces are instances of the same process.
\item For each action there exists a corresponding transition in the net that will be denoted, for simplicity, by the same name of the action.
\end{itemize}

\subsection{Log replay algorithm}\label{logreplayAlg}
The key algorithm exploited to analyze a Petri net model with respect to an event log is the {\itshape log replay} algorithm (see e.g.~\cite{DBLP:journals/is/RozinatA08}).
%DBLP:journals/widm/AalstAD12,
%,DBLP:conf/bpm/Munoz-GamaC10
Given a Petri net model and an event log as input to the algorithm, for each trace in the log, the algorithm starts by placing one token in the start place of the net. For each event in the trace the corresponding transition is fired assuming a  non-blocking behavior, then the marking of the net is updated. \emph{Non-blocking replay} means that whenever a firing of a disabled transition is needed, the algorithm enables the transition either by creating artificial tokens in the pre-set or, if possible, by firing some invisible transitions; the non-determinism in this procedure is resolved with a suitable cost function. For each trace, the result of the log replay execution is a map indicating which transitions of the model have been fired for each event in the trace, and additionally which tokens were artificially created and which remained  upon completion of the trace.
%% sequence on a trace can be represented as an ordered list $R$ of pairs $(tr, i)$, representing that the transition $tr$ has been fired to mimic the event $T[i]$.
The output of the log replay algorithm  can be used to check the conformance of the traces.
% and to evaluate some performance metrics.

% The log replay result can be used to evaluate both the conformance and performance of a business process.

\subsection{Conformance Analysis}\label{ConformanceAnalysis}
The goal of the conformance analysis is to check if a trace complies with the Petri net modeling the business process. Conformance problems can be discovered by analyzing tokens artificially created during the replay {\itshape (the missing tokens)} and tokens not consumed {\itshape (the remaining tokens)}. Figure~\ref{ConfLog}  presents two different traces, $T$ and $T'$, of the business process presented in Section~\ref{example}, where only the actions and the timestamps are reported. The log replay of trace $T$, which is compliant with the Petri net, terminates with a single remaining  
%marking containing one 
token in the end place \{${p13 \rightarrow 1}$\} and without reporting  missing tokens.

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.9\linewidth]
{./items/logConforme.pdf}
\caption{Conforming ($T$) and non-conforming ($T'$) traces}
\label{ConfLog}
\end{figure}

@@

\begin{comment}
The sequence returned by log replay is:
\begin{equation}
\begin{split}
R'&=\{(Order\_start,1), (Order\_complete,2), (WarehouseCheck\_start,3), \\
& (FinancialCheck\_start,4), (WarehouseCheck\_complete,5),(t1,6),(t2,6),\\
& (Notify\_start,6), (Notify\_complete,7)\}
\end{split}
\end{equation}
\end{comment}

\begin{comment}
\subsection{Performance analysis}\label{PerformanceAnalysis}
The performance of a process can be analyzed by exploiting the log
replay results. Since the logs contain timestamps, the log replay can
be used to compute performance measures of the process my measuring the elapsed time between production and consumption of tokens. Performance analysis can be applied only to traces that do not generate missing tokens during the replay since such tokens cannot have time information~\cite{DBLP:conf/wsfm/BruniCFFGS11} . During the log replay the following metrics can be computed for each trace and each place:
\begin{itemize}
\item sojourn time $(tsj)$ : the time interval between arrival and departure of tokens;
\item synchronization time $(tsc)$: the time interval between arrival of a token in the place and enabling of a transition in the post-set of the place;
\item waiting time $(tw)$:  the time interval between enabling of a transition in the post-set of the place and token departure (thus $tsj=tsc+tw $).
\end{itemize}

The duration of an activity is an important performance measure as well, and it is given by the waiting time of the only place between the start and the complete transition of the activity.
\end{comment}
\begin{comment}
To clarify the metrics evaluated by this technique, as done with conformance analysis, we exploit the Petri net  presented in Section~\ref{example} and the trace T presented in Figure~\ref{ConfLog}. The replay starts with firing the transition $Order\_start$ at time $0s$, so a token arrives at $p2$ at time $0s$. After $45 min$ the token in $p2$ is consumed and the transition $Order\_complete$ is fired, so $tw(p2)=45min$, and a token arrives at $p3$ and $p4$. At the time $1h:5min:25s$ the transition $WarhouseCheck\_start$ is fired, thus $tw(p4)=20min+25sec$ and a token is produced at $p6$. At the time $1h:10min:20s$ the transition $FinancialCheck\_start$ is fired and a token is produced at $p5$, so $tw(p3)=25min+20s$. After $2h:30min:20s$ from the replay start the transition $Financial\_complete$ is fired, hence we have $tw(p5)=1h+20min$ and a token is produced at $p8$. At time $3h:10min:20s$ the firing of the transition $WarehouseCheck\_complete$ is executed, so $tw(p6)=2h+4min+55s$ and a token is produced at $p7$. After $15 min$ the invisible transition $t1$ is fired and a token is produced at $p10$. At this point the transition $t2$ is enabled, so its firing is now possible and a token is produced at $p11$. Notice that $tsc(p8)=40min, tsc(p10)=0s$. At the time $3h:25min:20s$ the transition $Notify\_start$ is fired and a token is produced in $p12$, after $4min+4sec$ the transition $Notify\_complete$ is fired and $tw(p12) = 4min+40s$.

It is worth noting that the synchronization time is greater than zero
only for places which contain in their post-set transitions depending
from other places. For the Petri net in Figure~\ref{pnet}, the only
places which can have a non-null synchronization time are $p8$ and
$p10$. For example, the synchronization time is greater than zero for
$p8$ in the trace $T$ in Figure~\ref{ConfLog}, representing that, for
this particular process instance, the branch performing the financial
checking is faster than the warehouse branch.

\end{comment}

%\section{An approach based on classification for Process Analysis}\label{ClassifApproach}
\section{Classification for Process Analysis}\label{ClassifApproach}
Exploiting machine learning techniques for process analysis is encouraged by the presence of a huge amount of data recorded as attributes in event logs. The implicit information contained in those data could be significant for the process behavior analysis. In fact, this potential information could contain an explanation for the deviations discovered during conformance analysis, and for the performance level provided by the process. For this reason, we are interested in discovering how the process data may influence the conformance and the performance of the process executions. This is done with an approach based on \emph{classification}, a classical machine learning technique able to detect patterns on data in correspondence to which the process assumes a specific behavior.

\subsection{Classification: basic concepts}\label{classifBasicConcept}

@@

A classification model can be useful for both a descriptive and a predictive purpose. In fact it provides an explanatory tool to distinguish between objects of different classes, and at the same time it can be used to predict the class label of new records. In this way the model acts as a black box that automatically assigns a class label to an attribute set of a new record.

%purpose: it provides an explanatory tool to distinguish between objects of different classes, and for a predictive purpose: the model can be used to predict the class label of new records. In this way the model acts as a black box that automatically assigns a class label to an attribute set of a new record.

There is a general approach to solve a classification problem from a given dataset that is independent from the specific classification model. Each classification technique uses a learning algorithm to identify a model that best fits the relationship between the attribute sets and the class labels of the input data. The generated model should both fit the input data well, and correctly predict the class labels of new records. Finding the right trade-off between this two goals is the most delicate part of a classification technique, and in general of any machine learning task.

\begin{figure}[h]
\centering
\includegraphics[width=200pt]
{./items/classifApproach.pdf}
\caption{General approach for solving a classification problem.}
\label{figclassApproach}
\end{figure}

Figure~\ref{figclassApproach} summarizes the general approach for
solving a classification problem~\cite{tan2005introduction}. A \emph{training set}
consisting of records whose class labels are known must be provided to
the learning algorithm, and it is used to build the classification
model. This model is then applied to a \emph{test set}, which consists
of records with unknown class labels.
This step allows to  estimate the \emph{model accuracy} using one of
the standard approaches. For example, the accuracy can be represented by the
percentage 
%fraction of number the 
of records correctly predicted 
with respect to the total number of records in the test set.

Many classification models can be applied to a classification
problem. For our purpose we choose the \emph{decision tree} model
since the rules detected by the classification are shown explicitly,
providing an overview of how process data affects the
process behavior. Furthermore, this choice is encouraged by the simplicity of decision trees. In a decision tree, a class label (possible value of the target attribute) is assigned to each leaf node. The non-terminal nodes, which include the root and the internal nodes, contain attribute test conditions to filter records that have different characteristics.
\begin{comment}
In principle, there are exponentially many decision trees that can be constructed from a given set of attributes. While some of the trees are more accurate than others, finding the optimal tree is computationally infeasible because of the exponential size of the search space. However, efficient algorithms have been developed to induce a reasonably accurate decision tree in a reasonable amount of time. These algorithms usually employ a greedy strategy that grows a decision tree by making a series of decisions about which attribute to use for partitioning the data. For instance Hunt's algorithm is one of them and it represents the basis of many existing decision tree induction algorithms, including $C4.5$ algorithm used for our purposes.
\end{comment}
\subsection{Classification for Conformance Checking}\label{ClassConformance}
In order to identify the possible causes of non-conformance, a
classification problem can be formulated. The classification dataset
is extracted from the process data: for each process instance,
relevant information is  extracted and included as a record in the
dataset.
For the process presented in Section~\ref{example}, a record includes 
the following attributes: an instance identifier, a 
client identifier, the client typology (new or consolidated client),
the sales manager responsible for the order, the financial officer who
conducts the financial evaluation activities, the warehouseman
responsible for the warehouse checking, the supplying responsible name
in case of a provision, and finally the result communicated to the
client for the order issued. Furthemore, as target attribute 
we associate with each record the conformance result for that instance, 
which is a boolean value that is extracted from the output of the log replay.
% process
%instance provides an additional and important data, 
All these attributes are discrete and contribute to build the dataset
presented in Table~\ref{tab:SaleData} of the classification
problem. In this preliminary work, 
in order to 
%Notice that, since our goal is to 
explore the applicability
of existing machine learning  techniques to process analysis,
%rather then the
%evaluation of a new data mining algorithm, 
we just used synthetic data
generated to emulate process executions.

In order to build the classification problem, the attributes characterizing the
process are extracted from the log, while the conformance attribute is
computed using the conformance checking algorithm.
The Petri net of Figure~\ref{replayResult}  shows the
output of the conformance analysis executed on the event log:
edges are labeled with the number of times they were activated,
places are labeled with the number of remaining and missing tokens.
\begin{figure}[h]
\includegraphics[width=360pt]
{./items/Sales_PN_resultMod.pdf}
\caption{Petri net: conformance results.}
\label{replayResult}
\end{figure}

@@

Moreover, 37 remaining tokens in place $p3$ indicate the number of
times that the financial check activities were not performed.
Taking into account the application scenario, we can conclude that 
37 instances of the process are not compliant with the sale policy since
they did not execute the financial check activity.

% This situation is possible only if the invisible transition $t2$ is not enabled. Now, if $t2$ is not enabled that means that there are no tokens in the places $p8$ or $p10$. But notice that there are 21 remaining tokens in the place $p10$, so the transition $t2$ is not enabled for those $37$ instances because the place $p8$ does not have any token. Moreover, the Petri net in Figure~\ref{replayResult} shows 37 remaining tokens in place $p3$, so we can deduce that in those instances financial activities were not performed. From these facts, we can conclude that the conformance problems presented in  37 instances are due to a missing execution of the financial activity checking. On the other hand, the decision tree reports that the instances with conformance problems are made by consolidate clients.

\begin{table}[!h]
\scriptsize{
\centering
\begin{tabular}{|p{1cm}|p{1cm}|p{1,5cm}|p{1,2cm}|p{1,4cm}|p{1,2cm}|p{1,5cm}|p{1,4cm}|p{0,8cm}|}
\hline OrdIde & CltIde & CltType & SalMan & FinOff & WrhsMan & SupplyResp & OrdResut & Conf\\
\hline
1 & 20 & consolidate & Marco & Mary & Alex & Gianni & positive & no\\
\hline
2 & 15 & new & Anna & Johann & Roberto & Mario & positive & yes\\
\hline
3 & 10 &consolidate & Maria & Mary & Alessio & Gianni & negative & no\\
\hline
10 & 18 & consolidate & Johann & Mary & Roberto & Gianni & positive & yes \\
\hline
... & ... & ... & ... & ... & ... & ... & .... & ...  \\
\hline
\end{tabular}
}
\caption{Dataset for the conformance analysis.}
\label{tab:SaleData}
\end{table}
\normalsize

@@

The results of the conformance checking enable the use of 
existing machine learning  tools, allowing us to mine the decision tree of Figure~\ref{salesDecTree} as classification model for the sale process.
The resulting decision tree describes a data pattern in correspondence to which a process instance could present conformance errors: order managed by the sales manager \emph{Mary} and received from consolidate clients of the organization may not respect the standard sale procedure. 
%To get more significant information for our analysis,

@@

\begin{comment}
\item \texttt{Generate Instances with Performance}: the plug-in takes as input an event log, an indication of the synchronization point to examine and the performance result computed by the log replay. The output is a training set as presented in Table~\ref{tab:SaleDataPerf}.
\end{comment}

@@

\subsection{Methodology}
This last paragraph is dedicated to a description of the methodology to be followed to analyze a business process under the approach presented in this paper.

It should be clear that the analysis presented in the previous sections assumes the presence of a Petri net model of the business process to be analyzed. The model can be expressed in BPMN  as well, but in this case a transformation into a Petri net can be done through the techniques described in~\cite{DBLP:conf/wsfm/BruniCFFGS11}. The classification analysis  needs also an event log tracing the process behavior over multiple instances. Given a process model $M$ and an event log $L$, the methodology to be followed for the conformance analysis with the approach presented in Section~\ref{ClassConformance} can be summarized by these steps:
\begin{enumerate}

\item \label{step1} \textbf{Log replay analysis}: this phase consists of an execution of the log replay with  the model $M$ and the log $L$ as inputs. The results consist of two structures characterizing both the conformance and the performance of the traces in $L$. Let us call them \emph{ConfResult} and \emph{PerfResult}.

\item \label{step2} \textbf{Training set construction}: this phase aims at building the training set for the classification technique. This is done through an execution of the plug-in \texttt{Generate Instances with Conformance} with the event log $L$ and the conformance result \emph{ConfResult} as inputs. Let us call the obtained training set $TS$.

\item \textbf{Classifier building}: in this step the classification model (decision tree) for the conformance analysis is built using the plug-in \texttt{Generate Classifier} with the training set $TS$ as input. Let us call the obtained decision tree  $DT$.

\item \textbf{Using the classifier}: in this step the goal is to exploit $DT$ and the log replay result \emph{ConfResult} to establish the corrective measures in case of anomalies. Note that both $DT$ and \emph{ConfResult} can be visualized as in Figures~\ref{salesDecTree} and~\ref{replayResult} into ProM6 thanks to the plug-ins for resources managing that integrate the framework. $DT$ can also be used in a predictive way to establish the conformance of new process instances without needing a log replay execution. Given an event log containing new process instances, a dataset can be built through the plug-in \texttt{Generate Instances to Classify} and then classified by the plug-in \texttt{Classify Instances} according to the model $DT$.
\end{enumerate}

For a  performance analysis based on the extension summarized in Section~\ref{ClassPerf}, the same methodology can be followed modifying step~\ref{step2}. The construction of the training set is done through a plug-in (analogous to \texttt{Generate Instances with Conformance}) that takes as input \emph{PerfResult} obtained by step~\ref{step1} and an indication about the synchronization point chosen for the analysis.

>>>>

\documentclass{llncs}
\usepackage[latin1,utf8x]{inputenc}
\usepackage{color}
\usepackage[pdftex]{graphicx}
\usepackage{amsmath}
\usepackage{verbatim}
\usepackage{marginnote}

@@

\title{Classification Techniques for Conformance and Performance Checking in Process Analysis}
\author{Hind Chfouka\inst{1}, Andrea Corradini\inst{1} and Roberto Guanciale\inst{2}}
\institute{Department of Computer Science, University of Pisa, Italy \\ 
\email{\{chfouka,andrea\}@di.unipi.it}\and School of Computer Science and Communication,
KTH Royal Institute of Technology, Stockholm, Sweden \\
\email{robertog@kth.se}}	
\date{today}

@@

@@

\section{The Case Study: a Sale Business Process}\label{example}
A business process can be seen as a collection of activities occurring within an organization that lead to a specific goal. There are several formalisms for representing business processes. In this paper we represent them using Petri nets~\cite{DBLP:conf/ifip/Petri62,DBLP:journals/jcsc/Aalst98} since this makes possible to exploit some existing process analysis techniques that are based on such nets. However, in a business management context, Petri nets can be not expressive enough and usually process models are presented as workflows, for example using a standard notation for the business process modeling: BPMN (Business Process Model and Notation)~\cite{BPMN}. Transforming a BPMN model into a Petri net is possible thanks to a mapping technique presented in~\cite{DBLP:conf/wsfm/BruniCFFGS11}.
\begin{figure}[h]
\includegraphics[width=360pt]
{./items/Sales_PN.pdf}
\caption{Petri net for a sale process}
\label{pnet}
\end{figure}
%
Figure~\ref{pnet} presents the Petri net model for a sale
process.\footnote{\small{
The Petri net of Figure~\ref{pnet} is obtained by
transforming a BPMN model using the algorithm of~\cite{DBLP:conf/wsfm/BruniCFFGS11}, and mapping
each activity to a pair of start/end transitions. The resulting net
has been simplified by removing the unnecessary \emph{invisible
  transitions} used to encode the \emph{Join} or \emph{Fork} gateway~\cite{DBLP:conf/wsfm/BruniCFFGS11}.
}} 
This business process abstractly represents a fragment of the procedure followed in a commercial organization for managing orders received from clients. In general this procedure includes several activities and each of them involves a specific department of the organization. We assume that the sale process begins with the notification of an order from a client, so the first activity is simply called \emph{Order}. After the order is received, the sale process continues with some activities that can be done in parallel, since they do not present any dependency. The \emph{FinancialCheck} activity represents the financial analysis of the order (i.e verifying the financial situation of the client). Concurrently, the \emph{WarehouseCheck} activity checks the availability of merchandise requested by the order and, in case of a shortage, starts a supplying procedure. This is done through the activity \emph{ExternalWarehouse}. After the completion of the  parallel activities, a synchronization is needed before continuing the sale procedure with the final activity called \emph{Notify}, by which a result regarding the acceptance of the order is communicated to the client.

\section{Process Analysis}\label{Background}
Process analysis is performed using the Petri net representing the process and an event log recording the related executions, also called \emph{process instances}. The basic building blocks of event logs are \emph{events}, that record the execution of the process actions. Therefore an event $e$ can be seen as a tuple including the action recorded, a corresponding timestamp, and possibly other attributes describing relevant information about the state of the process. Events that belong to the same process instance are grouped in a {\itshape trace}, where events are ordered  by timestamp. An \emph{event log} is a set of traces, recording the activities performed by a system during a finite number of process executions.  In this paper the following assumptions are made:
\begin{itemize}
\item All traces are instances of the same process.
\item For each action there exists a corresponding transition in the net that will be denoted, for simplicity, by the same name of the action.
\end{itemize}

\subsection{Log replay algorithm}\label{logreplayAlg}
The key algorithm exploited to analyze a Petri net model with respect to an event log is the {\itshape log replay} algorithm (see e.g.~\cite{DBLP:journals/is/RozinatA08}). Given a Petri net model and an event log as input to the algorithm, for each trace in the log, the algorithm starts by placing one token in the start place of the net. For each event in the trace the corresponding transition is fired assuming a  non-blocking behavior, then the marking of the net is updated. \emph{Non-blocking replay} means that whenever a firing of a disabled transition is needed, the algorithm enables the transition either by creating artificial tokens in the pre-set or, if possible, by firing some invisible transitions; the non-determinism in this procedure is resolved with a suitable cost function. For each trace, the result of the log replay execution is a map indicating which transitions of the model have been fired for each event in the trace, and additionally which tokens were artificially created and which remained  upon completion of the trace. The output of the log replay algorithm  can be used to check the conformance of the traces.

\subsection{Conformance Analysis}\label{ConformanceAnalysis}
The goal of the conformance analysis is to check if a trace complies with the Petri net modeling the business process. Conformance problems can be discovered by analyzing tokens artificially created during the replay {\itshape (the missing tokens)} and tokens not consumed {\itshape (the remaining tokens)}. Figure~\ref{ConfLog}  presents two different traces, $T$ and $T'$, of the business process presented in Section~\ref{example}, where only the actions and the timestamps are reported. The log replay of trace $T$, which is compliant with the Petri net, terminates with a single remaining token in the end place \{${p13 \rightarrow 1}$\} and without reporting  missing tokens.

@@

\begin{figure}[h]
\centering
\includegraphics[width=345pt]
{./items/logConforme.pdf}
\caption{Conforming ($T$) and non-conforming ($T'$) traces}
\label{ConfLog}
\end{figure}

@@

\section{Classification for Process Analysis}\label{ClassifApproach}
Exploiting machine learning techniques for process analysis is encouraged by the presence of a huge amount of data recorded as attributes in event logs. The implicit information contained in those data could be significant for the process behavior analysis. In fact, this potential information could contain an explanation for the deviations discovered during conformance analysis, and for the performance level provided by the process. For this reason, we are interested in discovering how the process data may influence the conformance and the performance of the process executions. This is done with an approach based on \emph{classification}, a classical machine learning technique able to detect patterns on data in correspondence to which the process assumes a specific behavior.

\subsection{Classification: Basic Concepts}\label{classifBasicConcept}

@@

The general approach for solving a classification problem~\cite{tan2005introduction} consists of providing a \emph{training set} of records, whose class labels are already known, to the learning algorithm in order to build a classification model. This model is then applied to a \emph{test set}, which consists of records with unknown class labels. This step allows to  estimate the \emph{model accuracy} using one of the standard approaches. For example, the accuracy can be represented by the percentage  of records correctly predicted  with respect to the total number of records in the test set.

Many classification models can be applied to a classification problem. For our purpose we choose the \emph{decision tree} model since the rules detected by the classification are shown explicitly, providing an overview of how process data affects the process behavior. In a decision tree, a class label (possible value of the target attribute) is assigned to each leaf node. The non-terminal nodes, which include the root and the internal nodes, contain attribute test conditions to filter records that have different characteristics.

\subsection{Classification for Conformance Checking}\label{ClassConformance}
In order to identify the possible causes of non-conformance, a classification problem can be formulated. The classification dataset is extracted from the process data: for each process instance, relevant information is  extracted and included as a record in the dataset. For the process presented in Section~\ref{example}, a record includes  the following attributes: an instance identifier, a  client identifier, the client typology (new or consolidated client), the sales manager responsible for the order, the financial officer who conducts the financial evaluation activities, the warehouseman responsible for the warehouse checking, the supplying responsible name in case of a provision, and finally the result communicated to the client for the order issued. Furthemore, as target attribute we associate with each record the conformance result for that instance,  which is a boolean value that is extracted from the output of the log replay. All these attributes are discrete and contribute to build the dataset
presented in Table~\ref{tab:SaleData} of the classification problem. In this preliminary work,  in order to 
explore the applicability of existing machine learning  techniques to process analysis, we just used synthetic data
generated to emulate process executions. Note that for this  running example, attributes selected for the classification task are all focused on the actors of the process activities. In general, an attribute can characterize several aspects of an activity.

In order to build the classification problem, the attributes characterizing the process are extracted from the log, while the conformance attribute is computed using the conformance checking algorithm. The Petri net of Figure~\ref{replayResult}  shows the output of the conformance analysis executed on the event log: edges are labeled with the number of times they were activated, places are labeled with the number of remaining and missing tokens.

\begin{figure}[h]
\includegraphics[width=360pt]
{./items/Sales_PN_resultMod.pdf}
\caption{Petri net: conformance results.}
\label{replayResult}
\end{figure}

@@

Moreover, 37 remaining tokens in place $p3$ indicate the number of
times that the financial check activities were not performed.
Taking into account the application scenario, we can conclude that 
37 instances of the process are not compliant with the sale policy since
they did not execute the financial check activity.
{
\renewcommand{\arraystretch}{1.2}
\begin{table}[!h]
\scriptsize{
\caption{Dataset for the conformance analysis.}

\begin{tabular}{|p{1cm}|p{1cm}|p{1,5cm}|p{1,3cm}|p{1,3cm}|p{1,2cm}|p{1,5cm}|p{1,4cm}|p{0,8cm}|}
\hline OrdIde & CltIde & CltType & SalMan & FinOff & WrhsMan & SupplyResp & OrdResut & Conf\\
\hline
1 & 20 & consolidate & Marco & Mary & Alex & Gianni & positive & no\\
\hline
2 & 15 & new & Anna & Johann & Roberto & Mario & positive & yes\\
\hline
3 & 10 &consolidate & Maria & Mary & Alessio & Gianni & negative & no\\
\hline
10 & 18 & consolidate & Johann & Mary & Roberto & Gianni & positive & yes \\
\hline
... & ... & ... & ... & ... & ... & ... & .... & ...  \\
\hline
\end{tabular}
}
\label{tab:SaleData}
\end{table}
\normalsize

@@

The results of the conformance checking enable the use of 
existing machine learning  tools, allowing us to mine the decision tree of Figure~\ref{salesDecTree} as classification model for the sale process.
The resulting decision tree describes a data pattern in correspondence to which a process instance could present conformance errors: order managed by the sales manager \emph{Mary} and received from consolidate clients of the organization may not respect the standard sale procedure.

@@

@@

Figure \ref{dataflow} represents a data flow diagram that summarizes data dependencies between plug-ins discussed in this section.

\begin{figure}[h]
\includegraphics[width=360pt]
{./items/dataflow.pdf}
\caption{Data flow diagram}
\label{dataflow}
\end{figure}
